Scalable gradients of Gaussian Process models with celerite

### 2018-01-28 paper/ms.tex
angus citation and typo

This research note presents a derivation and implementation of efficient and
scalable gradient computation using the \emph{celerite} algorithm for Gaussian
Process (GP) modeling.
The algorithms are derived in a ``reverse accumulation'' or
``backpropagation'' framework and they can be easily integrated into existing
automatic differentiation frameworks to provide a scalable method for
evaluating the gradients of the GP likelihood with respect to all input
parameters.
The algorithm derived in this note uses less memory and is more efficient than
versions using automatic differentiation and the computational cost scales
linearly with the number of data points.

GPs \citep{Rasmussen:2006} are a class of models used
extensively in the astrophysical literature to model stochastic processes.
The applications are broad-ranging and some examples include the time domain
variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017}, data-driven models of light
curves or stellar spectra \citep{Wang:2012, Luger:2016, Czekala:2017}, and the
cosmic microwave background \citep{Bond:1987,Wandelt:2003}.
In all of these applications, the calculation and optimization of the GP
marginalized likelihood function (here we follow the notation of
\citealt{Foreman-Mackey:2017})
\begin{eqnarray}\eqlabel{loglike}
\log \mathcal{L}(\bvec{\theta},\,\bvec{\alpha}) &=&
    -\frac{1}{2}\,\left[\bvec{y} - \bvec{\mu}_\bvec{\theta}\right]^\T\,
        {K_\bvec{\alpha}}^{-1}\,\left[\bvec{y}-\bvec{\mu}_\bvec{\theta}\right]
    -\frac{1}{2}\,\log\det K_\bvec{\alpha} + \mathrm{constant}
\end{eqnarray}
is generally the computational bottleneck.
The details of these models are omitted here (see \citealt{Rasmussen:2006} and
\citealt{Foreman-Mackey:2017} for details), but the key point is that, for a
dataset with $N$ data points, every evaluation of a GP model requires
computation of the inverse\footnote{Or, more specifically, a linear operator
that can left multiply a vector or a matrix by the inverse of this matrix.}
and log-determinant of the $N \times N$ covariance matrix $K_\bvec{\alpha}$.
The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
general case, but the \emph{celerite} method was recently introduced in the
astronomical literature to compute the GP likelihood for a class of
one-dimensional models with $\mathcal{O}(N)$ scaling
\citep{Ambikasaran:2015, Foreman-Mackey:2017}.

>>>>

This research note presents a derivation and implementation of efficient and
scalable gradient computations using the \emph{celerite} algorithm for Gaussian
Process (GP) modeling.
The algorithms are derived in a ``reverse accumulation'' or
``backpropagation'' framework and they can be easily integrated into existing
automatic differentiation frameworks to provide a scalable method for
evaluating the gradients of the GP likelihood with respect to all input
parameters.
The algorithm derived in this note uses less memory and is more efficient than
versions using automatic differentiation and the computational cost scales
linearly with the number of data points.

GPs \citep{Rasmussen:2006} are a class of models used
extensively in the astrophysical literature to model stochastic processes.
The applications are broad-ranging and some examples include the time domain
variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017, Angus:2018}, data-driven
models of light curves or stellar spectra \citep{Wang:2012, Luger:2016,
Czekala:2017}, and the cosmic microwave background
\citep{Bond:1987,Wandelt:2003}.
In all of these applications, the calculation and optimization of the GP
marginalized likelihood function (here we follow the notation of
\citealt{Foreman-Mackey:2017})
\begin{eqnarray}\eqlabel{loglike}
\log \mathcal{L}(\bvec{\theta},\,\bvec{\alpha}) &=&
    -\frac{1}{2}\,\left[\bvec{y} - \bvec{\mu}_\bvec{\theta}\right]^\T\,
        {K_\bvec{\alpha}}^{-1}\,\left[\bvec{y}-\bvec{\mu}_\bvec{\theta}\right]
    -\frac{1}{2}\,\log\det K_\bvec{\alpha} + \mathrm{constant}
\end{eqnarray}
is generally the computational bottleneck.
The details of these models are omitted here (see \citealt{Rasmussen:2006} and
\citealt{Foreman-Mackey:2017} for details), but the key point is that, for a
dataset with $N$ data points, every evaluation of a GP model requires
computation of the inverse\footnote{Or, more specifically, a linear operator
that can left multiply a vector or a matrix by the inverse of this matrix.}
and log-determinant of the $N \times N$ covariance matrix $K_\bvec{\alpha}$.
The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
general case, but the \emph{celerite} method was recently introduced in the
astronomical literature to compute the GP likelihood for a class of
one-dimensional models with $\mathcal{O}(N)$ scaling
\citep{Ambikasaran:2015, Foreman-Mackey:2017}.
